---
title: 'My first pull request: imfr'
author: Mitsuo Shiota
date: '2020-10-28'
slug: my-first-pull-request-imfr
categories:
  - economics
tags:
  - R
output:
  blogdown::html_page:
    toc: false
---

## My first pull request

[Package 'imfr'](https://cran.r-project.org/web/packages/imfr/) is the first package which merged my very small pull request. Thank you, creator Christopher Gandrud. I would like to show a usage example below.

```{r libraries, results="hide", message=FALSE}
library(tidyverse)
library(imfr)
```

## Look for database ID

Package 'imfr' provides some functions to help us explore IMF database. First, I use 'imf_ids' function to get database IDs.

```{r get_database_ids, results="hide"}
id_list <- imf_ids()
```
```{r get_database_ids2}
nrow(id_list)
```

There are `r nrow(id_list)` databases. I get a dataframe with 2 columns, 'database_id' and 'description'.

```{r id_list}
head(id_list)
```

I transform 'description' into all lower characters, in order to make search easier.

```{r description_to_lower}
id_list <- id_list %>% 
  mutate(description = str_to_lower(description))
```

As I am interested in debt issues, I search 'debt' in 'description'. I get 'database_id' HPDD, of which 'description' is historical public debt.

```{r detect_debt}
id_list %>%
  filter(str_detect(description, "debt"))
```

## Retrieve code list of a database

I use 'imf_codelist' function to retrieve the code list of the database whose id is HPDD.

```{r get_codelist, results="hide"}
code_list <- imf_codelist(database_id = "HPDD")
```
```{r get_codelist2}
code_list
```

## Retrieve codes in each code list

Next, I use 'imf_codes' function to retrieve codes in a code list. As I use 'map' function from tidyverse for loop, I get 5 lists.

```{r get_codes, results="hide"}
meta <- map(code_list$codelist, imf_codes)
```

1st list is codes on scale.

```{r codes_scale}
meta[[1]] %>% head()
```

2nd list is codes on frequency.

```{r codes_frequency}
meta[[2]]
```

3rd list is codes on geographical areas. They are two-letter country codes defined in ISO 3166-1.

```{r codes_geographical_areas}
meta[[3]] %>% head()
```

4th list is codes on indicator. In this case, I get only one indicator GGXWDG_GDP, which describes Debt to GDP Ratio.

```{r codes_indicator}
meta[[4]]
```

5th list is codes on time format.

```{r codes_time_format}
meta[[5]]
```

## Download data

Now I am ready to download data. I use 'imf_data' function. I specify HPDD as 'database_id', and select 'indicator', 'country' and 'freq' in the code list. In this case, I don't select two-letter country codes, instead select "all" to get all available countries.

```{r download, results="hide"}
debt_gdp_ratio <- imf_data(database_id = "HPDD",
                           indicator = "GGXWDG_GDP",
                           country = "all",
                           freq = "A",
                           start = 1700, end = current_year()
)
```

I change 'year' from character to integer, and add 'country' by utilizing the 3rd list.

```{r make_pretty}
debt_gdp_ratio$year <- as.integer(debt_gdp_ratio$year)

debt_gdp_ratio <- debt_gdp_ratio %>% 
  left_join(meta[[3]], by = c("iso2c"="codes")) %>% 
  rename(country = description)

head(debt_gdp_ratio)
```

Some countries have data as old as 1800. They are U.K., U.S. and Sweden. Most recent year is 2015.

```{r year_range}
range(debt_gdp_ratio$year)
```

```{r countries_1800}
debt_gdp_ratio %>% 
  filter(year == 1800)
```

`r unique(debt_gdp_ratio$country) %>% length()` countries have some data.

## A chart and some thoughts

I use [countrycode.org](https://countrycode.org/) to get two-letter code of the select countries, and draw a chart.

Debt to GDP ratio rises in the war, and, if you win, it decreases thanks to growth and inflation. If you lose, it decreases thanks to some defaults and very high inflation. Some people in the losing side lose their financial wealth, but most misery does not come from debt, but from physical devastation.

Recently it rises due to secular stagnation, in which monetary policy is constrained by zero lower bound of nominal interest rates, and full employment can be achieved only by bubble or expansionary fiscal policy.

```{r line_chart}
debt_gdp_ratio %>% 
  filter(iso2c %in% c("US", "GB", "JP", "ZW")) %>% 
  ggplot(aes(year, GGXWDG_GDP, color = fct_reorder2(country, year, GGXWDG_GDP))) +
  geom_line(size = 1) +
  labs(y = "Debt to GDP Ratio",
       color = "counry")
```

Debt to GDP ratio reflects the government capacity to borrow, the household and the enterprise capacity to save, and the financial sector capacity to mediate. It melts, if nominal GDP growth rates are higher than nominal interest rates.

If Japan falls into deflation, nominal GDP growth rates are negative, and nominal interest rates are zero due to lower bound. This leads to ballooning debt to GDP ratios. Austerity is not the solution, as it leads to deeper deflation and under-employment. Coordinating and rebalancing the monetary and fiscal policies is the solution, which can make inflation expectation positive and well anchored.
