---
title: "Suntory Toku-cha"
author: "Mitsuo Shiota"
date: '2024-11-21'
categories:
  - R
knitr: 
  opts_chunk: 
    out.width: '70%'
---

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(tidytuesdayR)

theme_set(
  theme_light() +
  theme(
    legend.position = "none",
    plot.background = element_rect(fill = "darkgreen"),
    panel.background = element_rect(fill = "darkgreen"),
    panel.border = element_blank(),
    plot.title = element_text(color = "white", size = 16),
    axis.title = element_text(color = "white", size = 8),
    axis.text = element_text(color = "white"),
    axis.ticks.length = unit(-1, "mm"),
    axis.line = element_line(color = "white", linewidth = 0.3),
    axis.ticks = element_line(linewidth = 0.5),
    plot.caption = element_text(color = "white", size = 8),
    panel.grid = element_blank()
  )
)
```

## Suntory Toku-cha

Suntory is a beverage company. One of its products is Suntory Toku-cha. 'Toku' means food for specified health use, and 'Cha' means green tea. Suntory asserts this product helps reduce your body fat by showing a chart in [its advertisement](https://www.suntory.co.jp/softdrink/iyemon/tokucha/).

## Jpn Pharmacol Ther, vol. 40, no. 6, 2012

Ad chart is based on the research paper, 'Body Fat Reducing Effect and Safety Evaluation of Long-term Consumption of Green Tea Containing Quercetin Glucoside in Obese Subjects' published in Jpn Pharmacol Ther, vol. 40, no. 6, 2012.

Authors are six researchers of Suntory Wellness Ltd., and one doctor of a clinic.

They did RCT (Randomized Controlled Trial), and tested 36 outcomes related to abdomen fat area, body fat, and blood. 2 outcomes, abdomen fat area and blood HbA1c, are statistically significant, while other 34 outcomes, such as body fat ratio, are not significant. There is no mention about outcome pre-registration.

[Jpn Pharmacol Ther is reported to be an unreliable source because of its weak peer review](https://bunshun.jp/denshiban/articles/b9456).

```{r}
#| label: read_data

abdoment_fat_area <- tribble(
  ~group, ~week, ~mean, ~sd,
  "対照飲料", 0, 0, 0,
  "ケルセチン配糖体配合緑茶", 0, 0, 0,
  "対照飲料", 8, 5.45, 1.87,
  "ケルセチン配糖体配合緑茶", 8, -5.13, 1.99,
  "対照飲料", 12, 4.97, 2.34,
  "ケルセチン配糖体配合緑茶", 12, -5.32, 2.42
)

body_fat_ratio <- tribble(
  ~group, ~week, ~mean, ~sd,
  "対照飲料", 0, 0, 0,
  "ケルセチン配糖体配合緑茶", 0, 0, 0,
  "対照飲料", 4, 0.49, 0.28,
  "ケルセチン配糖体配合緑茶", 4, 0.32, 0.21,
  "対照飲料", 8, 0.50, 0.33,
  "ケルセチン配糖体配合緑茶", 8, 0.29, 0.26,
  "対照飲料", 12, 0.55, 0.33,
  "ケルセチン配糖体配合緑茶", 12, 0.21, 0.31,
)
```

## Suntory ad chart (replicated)

Suntory uses abdomen fat area to represent body fat.

```{r}
#| label: replicated_chart
#| fig-cap: Suntory ad chart (replicated) 
#| fig-align: center
#| fig-asp: 1

abdoment_fat_area |> 
  ggplot(aes(week, mean, color = group)) +
  geom_line(aes(linetype = group), linewidth = 1) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymax = mean + sd, ymin = mean - sd), width = 0.3) +
  annotate("text", x = 1, y = 5, label = "対照飲料", color = "white",
           hjust = 0) +
  annotate("text", x = 1, y = -5, label = "ケルセチン配糖体\n配合緑茶", color = "yellow", hjust = 0) +
  scale_color_manual(values = c("yellow", "white")) +
  scale_x_continuous(limits = c(0, 13), breaks = 0:3 * 4,
                     expand = expansion(add = c(0, 0))) +
  scale_y_continuous(limits = c(-10, 10), breaks = -2:2 * 5,
                     expand = expansion(add = c(0, 0))) +
  labs(x = "摂取期間（週）",
       y = "腹部全脂肪面積変化量 (cm2)",
       title = "＜腹部全脂肪面積変化量の推移＞",
       caption = "注：平均値 ± 標準偏差\n出典：薬理と治療40(6)495-503(2012)")
  
```

## Body fat ratio chart

Body fat ratio is the most commonly used measurement for body fat.

```{r}
#| label: body_fat_ratio_chart
#| fig-cap: Body fat ratio change
#| fig-align: center
#| fig-asp: 1

body_fat_ratio |> 
  mutate(week2 = week + if_else(week != 0 & group == "対照飲料", -0.2, 0)) |> 
  ggplot(aes(week2, mean, color = group)) +
  geom_line(aes(linetype = group), linewidth = 1) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymax = mean + sd, ymin = mean - sd), width = 0.3) +
  annotate("text", x = 1, y = 0.5, label = "対照飲料", color = "white",
           hjust = 0) +
  annotate("text", x = 1, y = -0.2, label = "ケルセチン配糖体\n配合緑茶", color = "yellow", hjust = 0) +
  scale_color_manual(values = c("yellow", "white")) +
  scale_x_continuous(limits = c(0, 13), breaks = 0:3 * 4,
                     expand = expansion(add = c(0, 0))) +
  scale_y_continuous(limits = c(-1, 1), breaks = -2:2 * 0.5,
                     expand = expansion(add = c(0, 0))) +
  labs(x = "摂取期間（週）",
       y = "体脂肪率変化（％）",
       title = "＜体脂肪率変化の推移＞",
       caption = "注：平均値 ± 標準偏差\n出典：薬理と治療40(6)495-503(2012)")
```

## Conclusion

[Takeshi Niinami, Representative Director, President & CEO, Suntory Holdings Limited, sees wider roles of the private sector for national health issues.](https://www.doyukai.or.jp/chairmansmsg/pressconf/2023/231003.html) If so, he should stop this ad.
